---
title: "R Notebook"
output: html_notebook
---

Collect Data
```{r}
# Load the library quantmod for financial time-series data collection
library(quantmod)

collect_data <- function(symbol, start_date, end_date) {
  getSymbols(symbol, src = "yahoo", from = as.Date(start_date), to = as.Date(end_date))
  
  closing_prices <- Cl(GSPC)
  
  return(closing_prices)
}
# COVID-19 data
closing_prices_19_23 <- collect_data("^GSPC", "2019-01-14", "2023-07-31")
```

```{r}
closing_prices <- closing_prices_19_23

# Plot the closing prices
plot(closing_prices, main="S&P 500 Daily Closing Prices (2019-2021)", ylab="Price ($)")
```

```{r}

directional_change_events <- function(prices, theta) {
  n <- length(prices)
  events <- rep("-", n)
  current_direction <- NA
  ext_price <- ext_idx <- NULL
  ext_price_high <- ext_price_low <- as.numeric(prices[1]) # Initialize with the first price
  ext_idx_high <- ext_idx_low <- 1

  for (i in 2:n) {
    # Get current price
    current_price <- as.numeric(prices[i])

    if (is.na(current_direction)) {
      # Track the highest and lowest prices seen
      if (current_price > ext_price_high) {
        ext_price_high <- current_price
        ext_idx_high <- i
      }
      if (current_price < ext_price_low) {
        ext_price_low <- current_price
        ext_idx_low <- i
      }

      # Check the magnitudes using both ext_prices and confirm the first DC event
      magnitude_high <- abs((current_price - ext_price_high) / ext_price_high)
      magnitude_low <- abs((current_price - ext_price_low) / ext_price_low)
      if (magnitude_high >= theta || magnitude_low >= theta) {
        if (magnitude_high > magnitude_low) { # Choosing the greater magnitude
          events[ext_idx_high] <- "EXT"
          current_direction <- "down"
        } else {
          events[ext_idx_low] <- "EXT"
          current_direction <- "up"
        }
        events[i] <- "DCC"
        ext_price <- current_price
        ext_idx <- i
        ext_price_high <- ext_price_low <- NULL
      }
    } else {
      
      # Calculate magnitude based on ext_price after the burn-in phase
      magnitude <- if (!is.null(ext_price)) abs((current_price - ext_price) / ext_price) else NA

      if (!is.null(ext_price) && magnitude >= theta) {
        if ((current_direction == "up" && current_price < ext_price) || (current_direction == "down" && current_price > ext_price)) {
          events[ext_idx] <- if (events[ext_idx] == "DCC") "DCC/EXT" else "EXT"
          events[i] <- "DCC"
          current_direction <- ifelse(current_direction == "down", "up", "down")
          #ext_price <- ext_idx <- NULL
          ext_price <- current_price # new
          ext_idx <- i # new
        }
      }

      if ((current_direction == "up" && current_price > ext_price) || (current_direction == "down" && current_price < ext_price)) {
        ext_price <- current_price
        ext_idx <- i
      }
    }
    
  }

  result <- data.frame(
    date = index(prices),
    price = as.numeric(prices),
    events = events
  )

  return(result)
}

```


```{r}
# Chen Ch 5 S&P 500 theta = 0.3%
theta <- 0.3 * 0.01

dc_events_summary <- directional_change_events(prices = closing_prices, theta)

head(dc_events_summary)
```

Plotting a similar figure to Chen Figure 2.1 - Use a small period of time. 

```{r}
# Create data frame
df <- data.frame(
  Time = 1:11,
  Price = c(97, 96, 98, 95, 100, 102, 104, 103, 105, 103, 99)
)
# Original annotations
original_anno_data <- data.frame(
  Time = c(4, 5, 9, 10, 11),
  Price = c(95, 100, 105, 103, 99),
  Labels = c('A', 'B', 'C', 'D', 'E')
)

# Additional annotations
new_anno_data <- data.frame(
  Time = c(4, 5, 9, 11),
  Price = c(95, 100, 105, 99),
  Labels = c('EXT', 'DCC', 'EXT', 'DCC')
)

# Create the plot with updates
plot_updated <- ggplot(df, aes(x = Time, y = Price)) +
  geom_line(color = "black") +
  geom_point(data = original_anno_data, aes(x = Time, y = Price), color = "black", size = 3) +
  geom_text(data = original_anno_data, aes(x = Time, y = Price, label = Labels), vjust = -1, color = "black") +

  # Additional annotations
  geom_text(data = new_anno_data, aes(x = Time, y = Price, label = Labels), vjust = 2, color = "black") +

  # Existing Horizontal blue and red lines
  geom_segment(aes(x = 4, xend = 5, y = df$Price[4], yend = df$Price[4]), linetype = "dotted", color = "blue") +
  geom_segment(data = df[4:5, ], aes(x = Time, xend = Time, yend = Price), y = df$Price[4], linetype = "dotted", color = "blue") +
  geom_segment(aes(x = 9, xend = 11, y = df$Price[9], yend = df$Price[9]), linetype = "dotted", color = "red") +
  geom_segment(aes(x = 11, xend = 11, y = df$Price[9], yend = df$Price[11]), linetype = "dotted", color = "red") +

  # New horizontal lines with annotations
  geom_segment(aes(x = 4, xend = 5, y = 106.25, yend = 106.25), 
               color = "lightblue", 
               arrow = arrow(type = "closed", ends = "both", length = unit(0.1, "inches"))) +
  geom_segment(aes(x = 5, xend = 9, y = 106.25, yend = 106.25), 
               color = "lightgreen",
               arrow = arrow(type = "closed", ends = "both", length = unit(0.1, "inches"))) +
  geom_segment(aes(x = 9, xend = 11, y = 106.25, yend = 106.25), 
               color = "lightblue",
               arrow = arrow(type = "closed", ends = "both", length = unit(0.1, "inches"))) +
  annotate("text", x = 4.5, y = 106, label = "DC Event", vjust = -1, 
           color = "lightblue") +
  annotate("text", x = 7, y = 106, label = "OS Event", vjust = -1, 
           color = "lightgreen") +
  annotate("text", x = 10, y = 106, label = "DC Event", vjust = -1, 
           color = "lightblue") +

  # Thetas and other configurations
  annotate("text", x = 11.2, y = df$Price[10] + 0.5, label = "θ", color = "red", size = 4) +
  annotate("text", x = 5.2, y = 97, label = "θ", color = "blue", size = 4) +
  scale_x_continuous(breaks = 1:11) +
  ylim(min(df$Price) - 2, max(df$Price) + 2) +
  labs(title = "", x = "Time", y = "Price ($)") +
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0.8)) +
  theme(axis.ticks = element_line(color = "black", size = 0.5))

# Display the updated plot
print(plot_updated)

```

```{r}
ggsave(filename = "dcevent.png", plot = plot_updated, width = 6, height = 4, dpi = 300)
```

DC ONGOING trend plot
```{r}
# Create data frame
df <- data.frame(
  Time = 1:11,
  Price = c(97, 96, 98, 95, 100, 102, 104, 103, 105, 103, 99)
)
# Original annotations
original_anno_data <- data.frame(
  Time = c(4, 5, 9, 10, 11),
  Price = c(95, 100, 105, 103, 99),
  Labels = c('A', 'B', 'C', 'D', 'E')
)

# Additional annotations
new_anno_data <- data.frame(
  Time = c(4, 5, 9, 11),
  Price = c(95, 100, 105, 99),
  Labels = c('EXT', 'DCC', 'EXT', 'DCC')
)

# Create the plot
plot <- ggplot(df, aes(x = Time, y = Price)) +
  geom_line(color = "black") +
  geom_point(data = original_anno_data, aes(x = Time, y = Price), color = "black", size = 3) +
  geom_text(data = original_anno_data, aes(x = Time, y = Price, label = Labels), vjust = -1, color = "black") +

  # Additional annotations
  geom_text(data = new_anno_data, aes(x = Time, y = Price, label = Labels), vjust = 2, color = "black") +

  # Horizontal blue line from point A to t=10
  geom_segment(aes(x = 4, xend = 10, y = df$Price[4], yend = df$Price[4]), linetype = "dotted", color = "blue") +
  
  # Vertical blue lines
  geom_segment(data = df[4:10, ], aes(x = Time, xend = Time, yend = Price), y = df$Price[4], 
               linetype = "dotted", color = "blue") +

  # Annotate "T" in blue
  annotate("text", x = 7, y = 94, label = "T", color = "blue") +

  # Annotate "TMV" on the side of one of the vertical blue lines
  annotate("text", x = 9.5, y = 98, label = "TMV", color = "blue") +

  # Horizontal red line from point C to above point E
  geom_segment(aes(x = 9, xend = 11, y = df$Price[9], yend = df$Price[9]), linetype = "dotted", color = "red") +
  
  # Vertical red line connecting to point E
  geom_segment(aes(x = 11, xend = 11, y = df$Price[9], yend = df$Price[11]), linetype = "dotted", color = "red") +

  # Annotate red theta beside the red vertical line
  annotate("text", x = 11.2, y = df$Price[10] + 0.5, label = "θ", color = "red", size = 4) +

  # Annotate blue theta beside the blue vertical line connecting to point B
  annotate("text", x = 5.2, y = 97, label = "θ", color = "blue", size = 4) +

  # Display every time point on the x-axis
  scale_x_continuous(breaks = 1:11) +
  
  # Increase y-axis limits
  ylim(min(df$Price) - 2, max(df$Price) + 2) +
  
  labs(title = "", x = "Time", y = "Price ($)") +
  theme_minimal() + 
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0.8)) +
  theme(axis.ticks = element_line(color = "black", size = 0.5))

# Display the plot
print(plot)


```
Save tracking TMV and T plot
```{r}
ggsave(filename = "ongoingTMV&Tplot.png", plot = plot, width = 6, height = 3, dpi = 300)
```

First, set up the DC event pairs - EXT(n) and DCC(n). 
```{r}
create_pairs <- function(dc_events_summary) {
  num_rows <- sum(dc_events_summary$events == "EXT") + sum(dc_events_summary$events == "DCC/EXT")
  pairs <- data.frame(date = as.Date(character(num_rows)), EXT = numeric(num_rows), DCC = numeric(num_rows))
  row_idx <- 1

  for (i in 1:nrow(dc_events_summary)) {
    event <- dc_events_summary$events[i]
    price <- dc_events_summary$price[i]
    date <- dc_events_summary$date[i]

    if (event == "EXT") {
      pairs[row_idx, "date"] <- date
      pairs[row_idx, "EXT"] <- price
    }

    if (event == "DCC") {
      pairs[row_idx, "DCC"] <- price
      row_idx <- row_idx + 1
    }
    
    if (event == "DCC/EXT") {
      pairs[row_idx, "DCC"] <- price
      pairs[row_idx + 1, "date"] <- date
      pairs[row_idx + 1, "EXT"] <- price
      row_idx <- row_idx + 1
    }
  }
  
  return(pairs)
}

# Creating the dc pairs data-frame
dc_pairs <- create_pairs(dc_events_summary)
print(dc_pairs)


```

Second, calculate DC indicators TMV and T.

Fix bizdays code tomorrow.
```{r}
if (!require("bizdays")) {
  install.packages("bizdays")
  library("bizdays")
}

# Define weekends
create.calendar(name="MyCalendar", weekdays=c("saturday", "sunday"))


# Create a function to calculate business days between successive dates
business_days_diff <- function(i) {
  if (i == 1) return(NA) # First row, no previous date to compare with
  bizdays(dc_pairs$date[i - 1], dc_pairs$date[i], "MyCalendar")
}
```


```{r}
dc_pairs <- dc_pairs %>%
  # TMV := total price movement of a DC event and an OS event
  mutate(TMV = abs((EXT - lag(EXT)) / (lag(EXT) * theta))) %>%
  
  # t := time for completion of trend
  mutate(t = as.numeric(difftime(date, lag(date), units = "days"))) %>%
  
  # tbiz := time for completion of trend (in business days)
  mutate(t_biz = sapply(1:n(), business_days_diff)) %>%

  # R := time-adjusted return of DC 
  mutate(R = theta * TMV / t) %>%
  
  # LR := the log transformation of the DC indicator
  mutate(LR = log(R))

dc_pairs
```

1. Fitting a hidden Markov model to our input data.
```{r, message=FALSE}

# Install and load the necessary package
if (!require("depmixS4")) {
  install.packages("depmixS4")
}

library(depmixS4)

# Prepare the data (depmixS4 requires data to be in data.frame format)
# Skip the first NA row
hmm_data <- data.frame(LR = dc_pairs[2:nrow(dc_pairs), 'LR'])

# Set seed
set.seed(48151623)

# Fit the model
hmm <- depmix(LR ~ 1, data = hmm_data, nstates = 2, family = gaussian())
fit <- fit(hmm)

# Print the fitted model
summary(fit)
```

```{r}
# Determine most likely state sequence
state_sequence <- viterbi(fit)

head(state_sequence)

# Combine closing_prices and state_sequence in a data frame
data_combined <- data.frame(date = dc_pairs$date[2:nrow(dc_pairs)],
                            LR = dc_pairs$LR[2:nrow(dc_pairs)], 
                            state = state_sequence$state)

head(data_combined)

# Subset the data based on state_sequence
data_state1 <- data_combined$LR[data_combined$state == 1]
data_state2 <- data_combined$LR[data_combined$state == 2]

# Now, data_state1 and data_state2 only contain closing prices corresponding to each state,
# without any NA values

# Calculate standard deviation for each state
sd_state1 <- sd(data_state1)
sd_state2 <- sd(data_state2)

# Print the standard deviations
print(paste0("Standard deviation for state 1: ", sd_state1))
print(paste0("Standard deviation for state 2: ", sd_state2))

# Switch states if needed
if (sd_state1 > sd_state2) {
  data_combined$state <- ifelse(data_combined$state == 1, 2, 1)
}

head(data_combined)

```
```{r}
data_combined
```

Next, we will produce Figure 5.5 which plots the detected regimes of the index. 
Then, we will summarise the time-series data using DC indicators, and then re-train the HMM.

```{r, warning=FALSE}
# Install ggplot2 if it's not already installed
if (!require("ggplot2")) {
  install.packages("ggplot2")
}

# Load the ggplot2 package
library(ggplot2)

# Install dplyr if it's not already installed
if (!require("dplyr")) {
  install.packages("dplyr")
}

# Load the dplyr package
library(dplyr)

p <- ggplot(data_combined, aes(x = date)) +
  geom_rect(aes(ymin = -8, ymax = 0,
                xmin = date, xmax = lead(date), fill = as.factor(state)),
            show.legend = T, alpha = 0.2) +
  geom_line(aes(y = data_combined$LR, colour = "Price")) +

  scale_fill_manual(name = "Regime", 
                    values = c("blue", "red"),
                    labels = c("State 1", "State 2")) +
  scale_colour_manual(name = "", values = c("black")) +
  
  scale_x_date(date_breaks = "1 year", date_labels = "%m/%y") +

  labs(x = "Date", y = "Log R",
       title = "") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) # Centering the title


# Print the plot
print(p)
```
```{r}
ggsave(filename = "hindsight_regimesPlot.png", plot = p, width = 6, height = 3, dpi = 300)
```

Formatting data in data_combined for table use
```{r}
data_combined2 <- data_combined
data_combined2$date <- format(data_combined2$date, "%d/%m/%Y")
```

Creating table to display regime periods
```{r}
create_regime_periods <- function(data_combined) {
  regimes <- data.frame(regime_1 = character(), regime_2 = character(), stringsAsFactors = FALSE)
  current_state <- data_combined$state[1]
  start_date <- data_combined$date[1]
  
  i <- 1

  # Special case when starting with regime 2
  if (current_state == 2) {
    regimes[nrow(regimes) + 1, "regime_1"] <- 'N/A'
    regimes[nrow(regimes), "regime_2"] <- paste(start_date, "to", data_combined$date[i])
    start_date <- data_combined$date[i + 1]
    current_state <- data_combined$state[i + 1]
    i <- i + 2
  }
  
  # Iterate through the data frame, checking for state changes
  while (i <= nrow(data_combined)) {
    if (current_state == 1) {
      while (i <= nrow(data_combined) && data_combined$state[i] == 1) {
        i <- i + 1
      }
      regimes[nrow(regimes) + 1, "regime_1"] <- paste(start_date, "to", data_combined$date[i - 1])
      start_date <- data_combined$date[i]
      current_state <- 2
    } else {
      while (i <= nrow(data_combined) && data_combined$state[i] == 2) {
        i <- i + 1
      }
      regimes[nrow(regimes), "regime_2"] <- paste(start_date, "to", data_combined$date[i - 1])
      start_date <- data_combined$date[i]
      current_state <- 1
    }
  }
  
  return(regimes)
}

# Create regime periods data frame
regime_periods <- create_regime_periods(data_combined2)
print(regime_periods)
```
Filling in gaps between regime periods
```{r}
# Parse dates
regime_periods2 <- regime_periods

regime_periods2$regime_1 <- strsplit(regime_periods2$regime_1, " to ")
regime_periods2$regime_2 <- strsplit(regime_periods2$regime_2, " to ")

# Convert to date
regime_periods2$regime_1 <- lapply(regime_periods2$regime_1, function(x) as.Date(x, format = "%d/%m/%Y"))
regime_periods2$regime_2 <- lapply(regime_periods2$regime_2, function(x) as.Date(x, format = "%d/%m/%Y"))

no_periods <- sum(!is.na(regime_periods$regime_1)) + sum(!is.na(regime_periods$regime_2))

for (i in 2:(no_periods)) {
  # when i even: compare regime 2 period start to previous regime 1 period end
  # and change regime 2 period start to fill in the gap 
  if (i %% 2 == 0) {
    regime_1_end <- regime_periods2$regime_1[[floor(i / 2)]][2]
    regime_2_start <- regime_periods2$regime_2[[floor(i / 2)]][1]
    
    if (as.numeric(difftime(regime_2_start, regime_1_end, units = "days")) != 1) {
      fixed_regime_2_start <- regime_1_end + 1
      regime_periods2$regime_2[[floor(i / 2)]][1] <- fixed_regime_2_start
    }
  } else {
    # when i odd: compare regime period 1 start to previous regime 2 period end
    regime_2_end <- regime_periods2$regime_2[[floor((i - 1) / 2)]][2] # i = 3 => i - 1 / 2 = 1, i = 5 => i - 1 / 2 = 2
    regime_1_start <- regime_periods2$regime_1[[floor((i + 1) / 2)]][1] # i = 3 => i + 1 / 2 = 2, i = 5 => i + 1 / 2 = 3
    
    if (as.numeric(difftime(regime_1_start, regime_2_end, units = "days")) != 1) {
      fixed_regime_1_start <- regime_2_end + 1
      regime_periods2$regime_1[[floor((i + 1) / 2)]][1] <- fixed_regime_1_start
    }
  }
}
```

Fixing regime_periods display of regime periods for the table in the report
Note - regime_periods2 contains the data for the periods in usable format rather than display format
```{r}
# Convert dates back to strings in the "DD/MM/YYYY" format
regime_periods$regime_1 <- lapply(regime_periods2$regime_1, function(x) format(x, "%d/%m/%Y"))
regime_periods$regime_2 <- lapply(regime_periods2$regime_2, function(x) format(x, "%d/%m/%Y"))

# Concatenate start and end dates into a single string separated by " to "
regime_periods$regime_1 <- sapply(regime_periods2$regime_1, function(x) paste(x[1], "to", x[2]))
regime_periods$regime_2 <- sapply(regime_periods2$regime_2, function(x) paste(x[1], "to", x[2]))

# Now regime_periods has been corrected for the missing dates in between detected regime periods
```

Produce LaTeX code for creating table of regime periods.
```{r}
if (!require("xtable")) {
  install.packages("xtable")
}
library(xtable)
regime_table <- xtable(regime_periods)
print(regime_table, type = "latex")
```
In the training data sets, the raw financial data is summarised into completed DC trends, under a threshold of 0.3%. The DC trends are then measured by two DC indicators: TMVEXT and TEXT.

```{r}
# Merging DC event data and HMM data (LR, state sequence)
merged_df <- merge(dc_pairs, data_combined, by = c("date", "LR"))
print(merged_df)

# arbitrary date chosen
DATE_SPLIT <- "2021-04-30"
train <- merged_df[merged_df$date <= DATE_SPLIT, ]
test <- merged_df[merged_df$date > DATE_SPLIT, ]
```

On the other hand, in the test data sets, the raw financial data was summarised into the on-going DC trends, which were then measured by two DC indicators TMV and T.

1. Filter result from directional_change_events function for test set dates
2. New function to track last known EXT price
3. New function to extract on-going pairs of EXT prices
4. Calculate TMV and T for on-going pairs of EXT prices

```{r}
test_dc_events_summary <- dc_events_summary[dc_events_summary$date >= (as.Date(DATE_SPLIT) + 1) & dc_events_summary$date <= index(tail(closing_prices, 1)), ]
rownames(test_dc_events_summary) <- 1:nrow(test_dc_events_summary)
```

Track last confirmed EXT price
```{r}
track_lastEXT <- function(test_dc_events_summary) {
  
  test_dc_events_summary$lastEXT <- NA
  test_dc_events_summary$lastEXTidx <- NA # new

  confirmDCC <- FALSE
  burnedin <- FALSE
  
  for (i in 1:nrow(test_dc_events_summary)) {
    
    if (!burnedin & test_dc_events_summary$events[i] == "EXT") {
      # not burned in and see first EXT => burned in
      burnedin <- TRUE
    }
    if (!burnedin) {
      next
    }
    
    # if see EXT => record price
    if (test_dc_events_summary$events[i] == "EXT") {
      lastEXTprice <- test_dc_events_summary$price[i]
      lastEXTidx <- i # new
    }
    
    # if see DCC => confirm last EXT price seen
    if (test_dc_events_summary$events[i] == "DCC") {
      confirmDCC <- TRUE
      newLastEXTprice <- lastEXTprice
      newLastEXTidx <- lastEXTidx # new
    }
    
    # if see DCC/EXT => see DCC and see EXT => confirm EXT price seen
    if (test_dc_events_summary$events[i] == "DCC/EXT") {
      confirmDCC <- TRUE
      # confirm last EXT price seen
      newLastEXTprice <- lastEXTprice
      newLastEXTidx <- lastEXTidx # new
      # record new EXT price
      lastEXTprice <- test_dc_events_summary$price[i]
      lastEXTidx <- i # new
    }
    
    # if DCC confirmed => record confirmed EXT
    if (confirmDCC) {
      test_dc_events_summary$lastEXT[i] <- newLastEXTprice
      
      test_dc_events_summary$lastEXTidx[i] <- newLastEXTidx # new
    }
    
  }
  
  return(test_dc_events_summary)
}

test_tracked_EXT <- track_lastEXT(test_dc_events_summary)
test_tracked_EXT$lastEXTdate <- test_tracked_EXT$date[test_tracked_EXT$lastEXTidx] # new
test_tracked_EXT
```

Calculate on-going TMV and T in test set
```{r}
test_tracked_TMV_T <- test_tracked_EXT %>%
  
  # calculate the value of tracked TMV in the on-going trend
  mutate(TMV = abs((price - lastEXT) / (lastEXT * theta))) %>%
  # calculate the value of tracked T in the on-going trend
  mutate(t = as.numeric(difftime(date, lastEXTdate, units = "days"))) %>%
  # calculate the value of tracked tbiz := time for on-going trend (in business days)
  
  # R := time-adjusted return of on-going DC 
  mutate(R = theta * TMV / t) %>%
  
  # LR := the log transformation of the DC indicator
  mutate(LR = log(R))

test_tracked_TMV_T
```

```{r}
# normalising train and test using min-max normalisation
min_max_normalisation <- function(df, column_name) {
  # Extract the column
  column <- df[[column_name]]

  # Compute min and max of the column
  min_value <- min(column, na.rm = TRUE)
  max_value <- max(column, na.rm = TRUE)

  # Perform min-max normalization
  normalised_column <- (column - min_value) / (max_value - min_value)

  # Add the normalized column to the data frame
  normalised_column_name <- paste(column_name, "normalised", sep = "_")
  df[[normalised_column_name]] <- normalised_column

  return(df)
}
```

Normalising the data
```{r}
# Normalising train data

# Normalize column 'TMV'
train_norm_TMV <- min_max_normalisation(train, 'TMV')

# Normalize column 't'
train_norm <- min_max_normalisation(train_norm_TMV, 't')

# Print the result
print(train_norm)


# Normalising test data

# Normalize column 'TMV'
test_norm_TMV <- min_max_normalisation(test_tracked_TMV_T, 'TMV')

# Normalize column 't'
test_norm <- min_max_normalisation(test_norm_TMV, 't')

# Print the result
print(test_norm)
```

In the empirical study, the parameters of the NBC are learnt from the training data sets. And the model is used to recognise market regimes for each pair of input features from the test data sets.

```{r}
# Subset relevant variables for NBC training
train_chen <- train_norm[, c('date', 'TMV_normalised', 't_normalised', 'state')]

test_chen <- test_norm[, c('date', 'TMV_normalised', 't_normalised')]

# LR and NORMALISING 

train_norm_LR <- min_max_normalisation(train, 'LR')
test_norm_LR <- min_max_normalisation(test_tracked_TMV_T, 'LR')

train_LR <- train_norm_LR[, c('date', 'LR_normalised', 'state')]

test_LR <- test_norm_LR[, c('date', 'LR_normalised')]

# CHOOSE DATA HERE
train_final <- train_chen
test_final <- test_chen
```

Need to take the state sequence from test (i.e. the state sequence over DC events found in hindsight by the HMM) and expand it over the whole period in test_final
Find final date in final regime period
```{r}
end_regime_date <- NA

if (is.na(regime_periods2[nrow(regime_periods2), 2][[1]])) {
  # if end in regime 1: get last date in final regime 1 detected
  end_regime_date <- regime_periods2[nrow(regime_periods2), 1][[1]][2]
} else {
  # else we end in regime 2: get last date in final regime 2 detected
  end_regime_date <- regime_periods2[nrow(regime_periods2), 2][[1]][2]
}

end_regime_date_idx <- as.numeric(rownames(test_final[test_final$date == end_regime_date, ]))
```

Adding state sequence to the whole period of dates in test set (i.e. expanding DC event state sequence)

```{r}
regime_periods2
```

```{r}
# Convert date column in test_final
test_final$date <- as.Date(test_final$date, format = "%d/%m/%Y") # changes date format

# Initialize state column
test_final$state <- NA

# For each row in test_final until the end regime last date, check if date falls within any regime period
for (i in 1:end_regime_date_idx) {
  for (j in 1:nrow(regime_periods2)) {
    # if date between regime 1 start and end dates
    if (test_final$date[i] >= regime_periods2$regime_1[[j]][1] & test_final$date[i] <= regime_periods2$regime_1[[j]][2]) {
      test_final$state[i] <- 1
      break
      # else if date between regime 2 start and end dates
    } else if (test_final$date[i] >= regime_periods2$regime_2[[j]][1] & test_final$date[i] <= regime_periods2$regime_2[[j]][2]) {
      test_final$state[i] <- 2
      break
    }
  }
}
```

```{r}
regime_periods
```

Algorithm 1 Naive Bayes Classifier

```{r}
# Naive Bayes Classifier

# Training Phase
train_NBC <- function(data) {
  # 1. Calculate the prior probability of each class
  priors <- table(data$state) / nrow(data)

  # 2. Calculate the mean and the standard deviation of the input features for each class
  summary_stats <- data %>%
    group_by(state) %>%
    summarise(mu_TMV = mean(TMV_normalised),
              sd_TMV = sd(TMV_normalised),
              mu_t = mean(t_normalised),
              sd_t = sd(t_normalised))

  # Return the model parameters
  list(priors = priors, summary_stats = summary_stats)
}

# Testing Phase
predict_class <- function(test_data, model) {
  # Apply the Gaussian distribution parameters to calculate the likelihood for each observation in the test data
  predicted_probs <- lapply(1:nrow(test_data), function(i) {
    # Get the TMV and t values for the current observation
    TMV <- test_data$TMV_normalised[i]
    t <- test_data$t_normalised[i]
    
    # If either TMV or t is NA, assign NA as the predicted class
    if (is.na(TMV) || is.na(t)) {
      return(list(p_x_given_Ck1 = NA, p_x_given_Ck2 = NA, posterior_k1 = NA, posterior_k2 = NA, pred_class = NA))
    }

    # For each class, calculate the probability p(x|Ck)
    likelihoods <- lapply(1:2, function(class) {
      mu_TMV <- model$summary_stats[model$summary_stats$state == class, ]$mu_TMV
      sd_TMV <- model$summary_stats[model$summary_stats$state == class, ]$sd_TMV
      mu_t <- model$summary_stats[model$summary_stats$state == class, ]$mu_t
      sd_t <- model$summary_stats[model$summary_stats$state == class, ]$sd_t

      p_TMV_given_class <- dnorm(TMV, mean = mu_TMV, sd = sd_TMV)
      p_t_given_class <- dnorm(t, mean = mu_t, sd = sd_t)

      # Use the Naive Bayes assumption to calculate the joint probability
      joint_probability <- p_TMV_given_class * p_t_given_class

      # Multiply by the prior probability of the class
      posterior_k <- joint_probability * model$priors[class]
      
      return(list(p_x_given_Ck = joint_probability, posterior_k = posterior_k))
    })

    # Compute total probability p(x)
    p_x = sum(sapply(likelihoods, function(x) x$posterior_k))
  
    # Compute posterior probability p(Ck|x)
    post_probs = sapply(likelihoods, function(x) x$posterior_k / p_x)
    
    # Predicted class is the one with the maximum posterior probability
    pred_class <- as.integer(which.max(post_probs))
    
    return(list(p_x_given_Ck1 = likelihoods[[1]]$p_x_given_Ck, 
                p_x_given_Ck2 = likelihoods[[2]]$p_x_given_Ck, 
                posterior_k1 = post_probs[1], 
                posterior_k2 = post_probs[2],
                pred_class = pred_class))
  })

  # Create a copy of test_data with additional columns
  test_data$p_x_given_Ck1 <- sapply(predicted_probs, function(x) x$p_x_given_Ck1)
  test_data$p_x_given_Ck2 <- sapply(predicted_probs, function(x) x$p_x_given_Ck2)
  test_data$posterior_k1 <- sapply(predicted_probs, function(x) x$posterior_k1)
  test_data$posterior_k2 <- sapply(predicted_probs, function(x) x$posterior_k2)
  test_data$pred_state <- sapply(predicted_probs, function(x) x$pred_class)

  return(test_data)
}

# Training the model
model <- train_NBC(train_final)

# Testing the model
predictions <- predict_class(test_final, model)
```

```{r}
predictions
```
Plot actual regimes detected by HMM
```{r}
predictions_plottable <- na.omit(predictions)
# Ensure state is treated as numeric
predictions_plottable$state <- as.numeric(as.character(predictions_plottable$state))
predictions_plottable$pred_state <- as.numeric(as.character(predictions_plottable$pred_state))

# Plotting
sandp_actual_regime_plot <- ggplot(predictions_plottable, aes(x = date, y = state, group = 1)) +
  geom_line(aes(group = 1)) + 
  labs(title = "S&P 500: Actual Regime", 
       x = "Date", 
       y = "State") +
  scale_x_date(labels = scales::date_format("%m/%Y"), date_breaks = "2 month") +
  scale_y_continuous(limits = c(1, 2), breaks = c(1,2)) + # Setting strict y-axis limits and breaks
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), aspect.ratio = 0.2) + # Adjust the aspect ratio here
  theme(plot.title = element_text(hjust = 0.5))
print(sandp_actual_regime_plot)
```
```{r}
ggsave(filename = "sandp_actual_regime_plot.png", plot = sandp_actual_regime_plot, width = 6, height = 2, dpi = 300)
```

```{r}
# Plotting
sandp_tracked_regime_plot <- ggplot(predictions_plottable, aes(x = date, y = pred_state, group = 1)) +
  geom_line(aes(group = 1)) + 
  labs(title = "S&P 500: Tracked Regime", 
       x = "Date", 
       y = "State") +
  scale_x_date(labels = scales::date_format("%m/%Y"), date_breaks = "2 month") +
  scale_y_continuous(limits = c(1, 2), breaks = c(1,2)) + # Setting strict y-axis limits and breaks
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), aspect.ratio = 0.2) + # Adjust the aspect ratio here
  theme(plot.title = element_text(hjust = 0.5))
print(sandp_tracked_regime_plot)
ggsave(filename = "sandp_tracked_regime_plot.png", plot = sandp_tracked_regime_plot, width = 6, height = 2, dpi = 300)

```

```{r}
predictions$state <- factor(predictions$state, levels=c(1, 2))
predictions$pred_state <- factor(predictions$pred_state, levels=c(1, 2))
```

```{r}
install.packages("caret")
library(caret)

# Create a confusion matrix
cm <- confusionMatrix(predictions$pred_state, predictions$state)

# Print confusion matrix
print(cm)

# Extract classification metrics
accuracy <- cm$overall['Accuracy']
sensitivity <- cm$byClass['Sensitivity']
specificity <- cm$byClass['Specificity']
precision <- cm$byClass['Pos Pred Value']
recall <- cm$byClass['Sensitivity'] # Recall is the same as sensitivity
F1 <- (2 * precision * recall) / (precision + recall) # Compute F1 score

# Print metrics
cat("Accuracy:", accuracy, "\n")
cat("Sensitivity:", sensitivity, "\n")
cat("Specificity:", specificity, "\n")
cat("Precision:", precision, "\n")
cat("Recall:", recall, "\n")
cat("F1 Score:", F1, "\n")

```









